---
title: "ps5-markdown"
author: "Noa Avitan"
date: "2023-02-17"
output: html_document
---

### 1: Load and check data (5pt)

Your first task is to do a very simple data check:

**1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title.**
See https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating.


**2. (2pt) Load data. How many rows/columns do we have?**

```{r message=FALSE, echo=FALSE}
library(tidyverse)
gap <- read_delim("gapminder.csv.bz2")
```
There are `r nrow(gap)` rows and `r ncol(gap)` columns in this data set.

**3. (2pt) Print a small sample of data. Does it look OK?**

```{r}
gap %>% 
  sample_n(5)
```

From what I can tell, the small sample of data looks okay (other than the missing columns).

### 2: Descriptive statistics (15pt)

**1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.**

```{r}
gap %>% 
  mutate(iso3_uni = n_distinct(iso3),
         iso2_uni = n_distinct(iso2),
         name_uni = n_distinct(name)) %>% 
  select(name_uni, iso3_uni, iso2_uni) %>% 
  head(1)
```


**2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?**

**(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?**

```{r message=FALSE}


```
Yes, there is one iso-2 code that corresponds to more name...

**(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries?**

_Hint: two of these entities are CHANISL and NLD CURACAO._


**3. (2pt) What is the minimum and maximum year in these data?**

```{r}
gap %>% 
  filter(!is.na(time)) %>% 
  summarise(min_year = min(time),
         max_year = max(time))
```


### 3: CO2 emissions (30pt)

Next, let’s analyze CO2 emissions.

**1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2 and co2_PC. Which years have most missing data?**

```{r message=FALSE}
gap %>% 
  group_by(time) %>% 
  summarise(num_co2 = sum(is.na(co2))) %>% 
  arrange(desc(num_co2)) %>% 
  head(3)

gap %>% 
  group_by(time) %>% 
  summarise(num_co2_PC = sum(is.na(co2_PC))) %>% 
  arrange(desc(num_co2_PC)) %>% 
  head(3)
```


**2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.**

```{r}
gap %>%
  group_by(name) %>% 
  filter(!is.na(co2)) %>% 
  summarise(co2_total = sum(co2)) %>% 
  filter(name == "China" | name == "United States of America" | name == "India" | name == "Israel" | name == "Morocco" | name == "Romania") %>% 
  ggplot(aes(name, co2_total)) +
  geom_col()
```

The U.S., China, and India are the top three countries in the graph with the highest amount of CO2 consumption. The additional countries I compared to, Israel, Morocco, and Romania, all have very minimal CO2 emissions in comparison.

**3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the same countries. What does this figure suggest?**

```{r}
gap %>%
  group_by(name) %>% 
  filter(!is.na(co2_PC)) %>% 
  summarise(co2_PC_total = sum(co2_PC)) %>% 
  filter(name == "China" | name == "United States of America" | name == "India" | name == "Israel" | name == "Morocco" | name == "Romania") %>% 
  ggplot(aes(name, co2_PC_total)) +
  geom_col()
```

This figure shows how much of the total consumption from each individual in the country. While the U.S. remains the country with the highest CO2 consumption per capita, Israel is actually the country with the 2nd highest amount (maybe because it has a smaller population). India and China both have relatively lower consumption rates compared to the total emissions, but it could be due to their large population.

**4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.**

_Note: Just compute averages over countries and ignore the fact that countries are of different size._
_Hint: Americas 2016 should be 4.80._

```{r}
gap %>%
  filter(time == 2016) %>% 
  group_by(region) %>% 
  filter(!is.na(co2_PC)) %>% 
  summarise(co2_PC_avg = mean(co2_PC))
```

Just for clarification purposes, in the question it didn't specify which year to refer to, but the hint referenced the Americas in 2016 so I decided to find the average based on 2016 only. There is an extreme outlier _NA_ in this data set which has an extremely high average, I'm honestly not sure why. The rest of the data seems reasonable considering their regions. I'm not surprised that Africa is the lowest, solely by understanding how majority of the countries are still trying to establish themselves.

#### 5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.


#### 6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).


### 4: GDP per capita (50pt)
Let’s look at GDP per capita (GDP_PC).

**1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better. Comment what do you see there.**


**2. (4pt) Make a similar plot, but this time use 2019 data only.**


**3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?**


**4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?**

_Note: here as average I mean just average over countries, ignore the fact that countries are of different size._


**5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.**

_Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up!_ 


**6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!**


**7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)**

_Hint: check out the function rank()!_
_Hint2: 17 for 1960._


**8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?**

_Hint: 0.0904 for 1960._


Finally, tell us how many hours did you spend on this PS.

I spent 3 hours on this PS.